Maximum Entropy Spectral Analysis: a case study

### 2021-05-05 paper/mesa.tex
a final pass through the paper

Finally, as already discussed Welch's method is very dependent on the choice of window function. 
Our choice, the Tukey window with aforementioned parameters, is the best compromise with found between noise 
and accuracy for the reconstruction, but different choices can be made, possibly providing more accurate results than the ones reported here. 
However, we want to stress that this fact does not invalidate our discussion but it reinforces it: one of the most appealing advantages of MESA is the minimal amount of fine tuning required.

@@

\section{Applications} \label{sec:applications}
\subsection{Temperature Time Series} \label{sec:temp_timeseries}[review]
We apply our code to the study of the temperature time series. This will allow to test our model performance on an example of great importance in physics. We will also assess the quality of the forecasting predictions: this task has a great impact also on practical applications.
We used the historical reanalysis data from the ``ERA5-Land hourly data from 1981 to present" dataset, downloaded from the Climate Data Store \cite{ERA5_dataset}.
Data refers to the temperature measure at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperature is measured once per hour for almost 31 years from 31st December 1989 to 30th November 2020.

In fig.~\ref{fig:temp_spectrum}, we compute the spectrum over all the data. We employ both FPE and CAT loss functions.
First of all, we note that at every frequency the estimation given by FPE is much less noisy the one from CAT: this issue was already studied in the previous sections. 
We clearly see that there is a peak on the PSD at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to a day. This is somehow trivial and highly expected.
Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they corresponds to signals that preserve the 1 day period while, at the same time, capturing the complex variability of the daily temperature cycle throughout the year.

Looking at the low frequencies, it is striking to note that FPE does not capture the peak $f_{yr} = \SI{1}{yr^{-1}}$ located at the frequency of a year. On the other hand, the peak is captured well by the CAT loss function.
In analogy as what observed for the peak at $f_D$, in the CAT spectrum we observe a higher order harmonic at a frequency $2 \cdot f_{yr}$. As in the previous case, this is required to better model the temperature variation in the year.

The failure of FPE in catching the low frequency trend can be understood by looking at fig.~\ref{fig:p_vs_ptrue}. It is shown that CAT systematically select very long filters (i.e. large $p_{CAT}$ values for the autoregressive orders), whereas FPE tends to select smaller values for the order $p_{FPE}$ of the autoregressive process, often understimating badly the actual value.
In this example, we have $p_{CAT} \sim  36500 \sim  \SI{4}{yr}$, whereas $p_{FPE} \sim  1300 \sim  \SI{2}{months}$.
Clearly, the autoregressive process selected by FPE cannot produce a meaningful prediction on the timescale longer than a few weeks and for this reason it is unable to capture the peak at $f_{yr}$. In other words, the model is too simple to model \textit{both} the behaviour at high and low frequencies (separated by approximately 3 orders of magnitude).
On the other hand, the model chosen by CAT is much more complex and is able to model also the high frequency behaviour.
The ``actual" length of the autoregressive process should be closer to the choice of CAT.

We now assess the accuracy of the forecasting of new observations of the time series.
Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ indepedent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temparature values.
The predictions spans a two years range of time.
We report our results in fig.~\ref{fig:temp_spectrum}.

@@

\subsection{Forecasting the LIGO strain} \label{sec:LIGO_forecasting}
As a last example of an application of MESA on real data, we go back to the strain produced by the LIGO-Virgo interferometers and we forecast the future observations.
We employ the data released to the public by the LIGO/Virgo collaboration \cite{Abbott_2021}.,
The spectral analysis is performed with CAT and FPE loss functions on $\SI{1000}{s}$ of data \cite{Abbott_2021} sampled at $\SI{4096}{Hz}$ observed at Livingston observatory starting from GPS time $1164603392$.
We then forecast the following $\SI{100}{s}$ of observations with the model optimized with CAT. The results are shown in fig.~\ref{fig:LIGO_forecast}.

By looking at the figure, we note that the forecasted prediction is always in the $90\%$ confidence interval.
The confidence interval, however, is very broad and the prediction is not very accurate (being the same order of magnitude of the strain).
By looking at the standard deviation $\sigma$ of the predictions (bottom panel in fig.~\ref{fig:LIGO_forecast}), we note that it increases very quickly in the first $\SI{0.5}{s}$ and then its trend stabilizes to a constant value.
The order of the autoregressive process selected is $p_{CAT} = 57766 \sim \SI{14}{s}$, which is much larger than the region in which $\sigma$ is small $\sim \SI{0.2}{s}$. This means that the series is very difficult to predict: the knowledge of past observations is not very helpful in predicting future observations.
Similar performances in $\sigma$ are observed also with the FPE model: in this case FPE selects a large prediction filter $p_{FPE} = 29924$ , similar to those selected by CAT.

Despite poor predictions on long timescales, the autoregressive process obtained with MESA still can be useful on short timescales. Indeed, a precise prediction of the strain time series can be beneficial in anomaly detection and removal of glitches. The predictions can form an expected baseline for the strain; any anomaly (i.e. a glitch or even a transient of physical origin) can show up as a large departure from expected trend.
Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subracting the expected signal with the actual signal.
This can work: a typical glitch \cite{Zevin_2017} can last as long as $\SI{0.2}{s}$, which is close to the time scale in which the forecasting is realiable.

As noted above, for pushing further the prediction performance from our simple linear predictor, more sophisticated forecasting methods are available in the Machine Learning literature (see e.g. \cite{LSTM_article, oord2016wavenet}) and they can be trained for precision forecasting. However, a Maximum-Entropy trained autoregressive process can be a simple baseline for comparing such more advanced methods and might suffice for many purposes.
This opens a promising path in GW data analysis and other fields of physics might also take advantage from this.

@@

\section{Final remarks and future prospects} \label{sec:conclusion}
In this work, we presented a case study of the application of Maximum Entropy principle in the realm of spectral estimation. Albeit the methodology hereby presented is grounded on solid theoretical foundations and its merits are widely recognised, Maximum Entropy methods have yet to be adopted routinely in the study of problems related to time series. The superior nature of maximum entropy methods and in particular of Burg's method is exemplified by the closed form estimate of the power spectral density and by the theoretical bridge it lies between spectral analysis and autoregressive processes. Moreover, the method presents, in our view, two main advantages when compared with more traditional ones; first there is no need to choose an arbitrary window function to correct the data and second it provides as straightforward way to compute predictions given past observations. Accompanying this work, we provide a publicly available \texttt{Python} implementation, called \texttt{memspectrum}, that we used to perform numerical studies of its behaviour and to  demonstrate its reliability. 
Since the order of the autoregressive process is not determined by the theory, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance than Welch algorithm. The use of MESA is particularly useful for short time series samples, where Welch method is outperformed in accuracy and confidence. 
This observation suggests a promising avenue to pursue in future developments: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD, hence MESA could be employed for a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalising over the detector noise PSD. Whenever the likelihood of the model for a time series takes the form
\begin{equation}
	\mathcal{L}(d_t | \theta) \propto exp\left[ -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2}{S(f)} \right]
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$. The typical approach is to estimate the PSD offline on a large batch of data and off-source (see~\cite{lalinference}), that might not reflect the structure of noise in the time slice to analyse. MESA can offer a way out from this problem: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. Preliminary studies suggest that this is possible~\cite{martini_thesis}.

Furthermore, MESA provides a simple, but robust and quite accurate in the short term, predictor for the time series. This fact is remarkable and can be used in time series analysis for several purposes. As discussed in Sec.~\ref{sec:LIGO_forecasting}, an anomaly detection pipeline can be easily built using the forecasts of MESA: the predictions can form a baseline to compare the actual observations with. Whenever the observed data are outside the expectations, an anomaly detection can be claimed. Of course such predictions can be done with a more accurate (perhaps nonlinear model); however MESA has the advantage of being simple and fast to train, while providing good (short term) predictions. At the same time, several instruments present gaps in their data stream, for instance LISA is expected to show such gaps~(e.g. \cite{lisa_gaps} and references therein), MESA forecasting capabilities could be used to fill those gaps with predicted data from past observations.

>>>>

Finally, as already discussed Welch's method is very dependent on the choice of window function. 
Our choice, a Tukey window with aforementioned parameters, is the best compromise with found between noise 
and accuracy for the reconstruction, but different choices can be made, possibly providing more accurate results than the ones reported here. 
However, we want to stress that this fact does not invalidate our discussion but it reinforces it: one of the most appealing advantages of MESA is the minimal amount of fine tuning required.

@@

\section{Applications} \label{sec:applications}
\subsection{Temperature Time Series} \label{sec:temp_timeseries}
As a further example of the breadth of applicability of MESA, we applied our implementation to atmospheric temperature time series. 
The reason for this choice is twofold: i) atmospheric temperature time series present a variety of overlapping periodicities most of which are 
known; ii) it provides a stress test for the time series forecast analysis.  As dataset, we  used the historical reanalysis data 
from the ``ERA5-Land hourly data from 1981 to present" dataset, downloaded from the Climate Data Store~\cite{ERA5_dataset}.
The data consist in temperatures taken at coordinates $\textrm{N}\;\ang{45;5;} \;\; \textrm{E}\;\ang{9;1;}}$, corresponding about to the city of Milan, Italy. The temperatures are given on an hourly cadence for almost 31 years from 31st December 1989 to 30th November 2020.

Fig.~\ref{fig:temp_spectrum}, shows the MESA spectrum inferred from the data. As a comparison, we adopt both the FPE and the CAT loss functions.
In agreement with what we found in previous sections, the FPE PSD is more regular compared to the one from CAT. Both spectra show a
peak at the frequency $f_D = \SI{1}{day^{-1}}$ corresponding to one day, corresponding to the day-night cycle. This is expected.
Higher order harmonics (corresponding to integer multiples of $f_D$) are also visible up to the Nyquist frequency ($f_D = \SI{0.5}{hour^{-1}}$): they corresponds to signals that preserve the 1 day period while, at the same time, capturing the complex variability of the daily temperature cycle throughout the year.

Looking at low frequencies, FPE does not capture the yearly variability at $f_{yr} = \SI{1}{yr^{-1}}$. On the other hand, the peak is captured well by the CAT loss function. In analogy as what observed for the peak at $f_D$, in the CAT spectrum we observe a higher order harmonics at a frequency $2 \cdot f_{yr}$. As in the previous case, this is required to better model the temperature variation in the year.

The failure of FPE in capturign the low frequency trend can be understood by looking at Fig.~\ref{fig:p_vs_ptrue}. It is shown that CAT systematically select very long filters (i.e. large $p_{CAT}$ values for the autoregressive orders), whereas FPE tends to select smaller values for the order $p_{FPE}$ of the autoregressive process, often underestimating the actual value.
In this example, we have $p_{CAT} \sim  36500 \sim  \SI{4}{yr}$, whereas $p_{FPE} \sim  1300 \sim  \SI{2}{months}$.
The autoregressive process selected by FPE, hence cannot produce a meaningful prediction on the timescale longer than a few weeks and for this reason it is unable to capture the low frequency peak at $f_{yr}$. In other words, the model is too simple to model \textit{both} the behaviour at high and low frequencies (separated by approximately 3 orders of magnitude).
On the other hand, the model chosen by CAT is much more complex and is able to model also the high frequency behaviour.
The ``actual" length of the autoregressive process should be closer to the choice of CAT.

We now assess the accuracy of the forecasting of new observations of the time series. 
Based on actual data, we try to predict future values as described in sec.~\ref{sec:forecasting} (of course the MESA is performed with CAT loss function). We produce $N =100$ indepedent predictions and we compute the median as well as the $90\%$ confidence interval. This is compared with the actual measured temparature values. The predictions spans a two years range of time. We report our results in Fig.~\ref{fig:temp_spectrum}.

@@

\subsection{Forecasting the LIGO strain} \label{sec:LIGO_forecasting}
As a last example of an application of MESA, we return to the study of the strain produced by the LIGO-Virgo interferometers 
and we forecast the future observations. We focus on the public data released by the LIGO/Virgo collaboration~\cite{Abbott_2021}.
We reconstruct the PSD both assuming the CAT and FPE loss functions on $\SI{1000}{s}$ of data \cite{Abbott_2021} from the Livingston observatory
starting from GPS time $1164603392$. The data are sampled at $\SI{4096}{Hz}$.

We then forecast the following $\SI{100}{s}$ of observations with the model optimized with CAT. The results are shown in fig.~\ref{fig:LIGO_forecast}.
The prediction is always in the $90\%$ confidence interval. The confidence interval, however, is very broad and the prediction is not very accurate (being the same order of magnitude of the strain).
By looking at the standard deviation $\sigma$ of the predictions (bottom panel in Fig.~\ref{fig:LIGO_forecast}), we note that it increases very quickly in the first $\SI{0.5}{s}$ and then it converges to a constant value.
The order of the autoregressive process selected is $p_{CAT} = 57766 \sim \SI{14}{s}$, which is much larger than the region in which $\sigma$ is small $\sim \SI{0.2}{s}$. The implication is that the series is very difficult to predict: the knowledge of past observations is not very helpful to predict future observations. Although FPE selects a prediction filter order $p_{FPE} = 29924$ smaller than CAT,  the forecasted time series behaves very similarly.

Despite poor predictions on long timescales, the AR process obtained with MESA still can be useful on short timescales. Indeed, a precise prediction of the strain time series can be beneficial in the detection of anomalies in the data and, eventually, their removal. The predictions can form an expected baseline for the strain; any anomaly (i.e. a glitch or even a transient of physical origin) can show up as a large departure from expected trend.
Moreover, if a glitch is detected, its shape can be estimated (as well as the confidence level) by subracting the expected signal with the actual signal.
This can work: a typical glitch~\cite{Zevin_2017} can last as long as $\SI{0.2}{s}$, close to the time scale over which the forecasting is reliable.

As noted above, for pushing further the prediction performance from our simple linear predictor, more sophisticated forecasting methods are available in the Machine Learning literature (see e.g. \cite{LSTM_article, oord2016wavenet}) and they can be trained for precision forecasting. However, a Maximum-Entropy trained AR process can be a simple baseline for comparing such more advanced methods and might suffice for many purposes.
This opens a promising path in GW data analysis and other fields of physics might also take advantage from this.

@@

\section{Final remarks and future prospects} \label{sec:conclusion}
We presented a case study of the application of Maximum Entropy principle to the realm of spectral estimation. Albeit the methodology hereby presented is grounded on solid theoretical foundations and its merits are widely recognised, Maximum Entropy methods have yet to be adopted routinely in the study of problems related to time series. The superior nature of maximum entropy methods, and in particular of Burg's method, is exemplified by the closed form estimate of the power spectral density and by the theoretical bridge between spectral analysis and AR processes. Moreover, the method presents, in our view, two main advantages when compared with more traditional ones; first there is no need to choose an arbitrary window function to correct the data and, second it provides as straightforward way to compute predictions given past observations. Accompanying this work, we provide a publicly available \texttt{Python} implementation, called \texttt{memspectrum}, that we used to perform the numerical studies presented in this work..

Since the order of the AR process is not yet determined by the theory, we opted for an in-depth investigation of several proposals in the literature and found that different loss functions are required for different situations, with the FPE loss function being the most indicated to deal with gravitational wave data. Along these lines, we directly compared the PSDs computed with MESA with the canonical Welch's algorithm. As outlined in Sec.~\ref{sec:Welch_comparison}, MESA provides PSD estimates with smaller variance than Welch algorithm. The use of MESA is particularly useful for short time series samples, where Welch's method is outperformed in accuracy and confidence. 
This observation suggests a promising avenue to pursue in future developments: for short time series, comparable with the length of binary black hole systems as observed by LIGO, Virgo and KAGRA, the computational cost of MESA is moderate and the inferred PSD is an accurate representation of the true underlying PSD, hence MESA could be employed for a simultaneous inference of the PSD and of the gravitational wave signal during a Bayesian parameter estimation exercise, effectively marginalising over the detector noise PSD. Whenever the log-likelihood of the model for a time series takes the form
\begin{equation}
	\log\mathcal{L}(d_t | \theta) \propto  -\frac{1}{2} \frac{(\tilde{d}(f)- \tilde{x}(f;\theta))^2 - \frac{1}{2}\log\left[\int  df\,S(f)} \right]}
\end{equation}
where $\tilde{\hphantom{a}}$ denotes the Fourier transform and the signal model $x(t;\theta)$, dependent on some parameter $\theta$, is a prediction for a deterministic signal buried in the observed time series $d(t)$. The typical approach is to estimate the PSD offline on a large batch of data and off-source (see~\cite{lalinference}), that might not reflect the structure of noise in the time slice to analyse. MESA can offer a way out from this problem: at each evaluation of the likelihood, a new spectral analysis is performed by computing the PSD on $d_t-x_t$.
In this way, the PSD would depend also on $\theta$ and would effectively model the residuals. Preliminary studies suggest that this is possible~\cite{martini_thesis}.

Furthermore, MESA provides a simple, but robust and quite accurate, albeit for short times, predictor for the time series. This fact is remarkable and can be used in time series analysis for several purposes. As discussed in Sec.~\ref{sec:LIGO_forecasting}, an anomaly detection pipeline could be built using the forecasts of MESA: the predictions can form a baseline to compare the actual observations with. Whenever the observed data are outside the expectations, an anomaly detection can be claimed. Of course such predictions can be done with a more accurate (perhaps nonlinear model); however MESA has the advantage of being simple and fast to construct, while providing good predictions. At the same time, several instruments present gaps in their data stream, for instance LISA is expected to show such gaps~(e.g. \cite{lisa_gaps} and references therein), MESA forecasting capabilities could be used to fill those gaps with predicted data from past observations.
